---
title: 'Markdown Document for Replicability Example'
author: "Douglas Luke"
date: "May 6, 2025"
output:
  pdf_document:
    toc: yes
  html_document:
    code_folding: show
    highlight: pygments
    toc: yes
    toc_float:
      collapsed: no
---


# The Network Analysis 'Five-Number Summary'

```{r chapfivesetup,message=FALSE,warning=FALSE}
library(igraph)
library(UserNetR2)
options(width=50, digits=3)
```

## Network Analysis in R: Where to Start

How should you start when you want to do a network analysis in R? The
answer to this question rests of course on the analytic questions you
hope to answer, the state of the network data that you have available,
and the intended audience(s) for the results of this work. The good
news about performing network analysis in R is that, as will be seen
in subsequent chapters, R provides a multitude of available network
analysis options. However, it can be daunting to know exactly where
to start.

In 1977, John Tukey introduced the five-number summary as a simple
and quick way to summarize the most important characteristics of
a univariate distribution. Networks are more
complicated than single
variables, but it is also possible to explore a set of important
characteristics of a social network using a small number of procedures in
R.

In this chapter, we will focus on two initial steps that are almost always
useful for beginning a network analysis: simple visualization,
and basic description using a 'five-number summary.'
This chapter also serves as a gentle
introduction to basic network analysis in R, and demonstrates how
quickly this can be done.

## Preparation


Similar to most types of statistical analyses using R, the first
steps are to load appropriate packages (installing them first if
necessary), and then making data available for the analyses. The
\texttt{igraph} network analysis package will be used
here and through much of the book for the analyses. The data used in this book and chapter are from 
the \texttt{UserNetR2} package
that accompanies the book. (See the previous chapter for instructions on how to download and 
install the data package.) The specific dataset used here is called
\texttt{Moreno}, and contains a friendship network of fourth
grade students first collected by Jacob Moreno in the 1930s.


To access an existing network data object in R, we would first load the appropriate
network analysis package (i.e., \texttt{igraph}) as well as the
data package itself (i.e., \texttt{UserNetR2}). (These steps will
not always be shown in the subsequent chapters, but these
need to be done before embarking on new analyses.) Once we 
have the packages loaded, then we will have access to the
network data file(s). We can use the ubiquitous \texttt{summary()}
function to make sure that R (and igraph) can see the
data object and know how to use it.

```{r tstMorenoSummary}
library(igraph)
library(UserNetR2)
summary(Moreno)
```

This shows that R recognizes Moreno as an igraph network object. 
We will explain some of these summary elements in Chapter \ref{chap:data_manage}
when we start looking at how to manage network data in R.

## Simple Visualization

The first step in network analysis is often to just take a
look at the network. Network visualization is critical, but as
Chapters \ref{chap:plotbasic} through \ref{chap:plotadv} indicate,
effective network graphics take careful planning and execution
to produce. That being
said, an informative network plot can be produced with one
simple function call. The only added complexity here is
that we are using information about the network members'
gender to color code the nodes. The syntax details
underlying this example will be covered in greater depth
in chapters \ref{chap:data_manage}, \ref{chap:plotbasic}, and
\ref{chap:graphicdesign}.


```{r plotmorenotext, eval=FALSE}
plot(Moreno,
     vertex.color = V(Moreno)$gender,
     vertex.size = 15)
```


```{r plotmoreno, fig.lp="fig:", fig.cap = 'Friendship ties among a 4th grade class, from Moreno',out.width='80%', fig.asp=.75, fig.align='center',echo=FALSE}
op <- par(mar=c(0,0,0,0))
plot(Moreno,
     vertex.color = V(Moreno)$gender,
     vertex.size = 15)
par(op)
```


The resulting plot makes it immediately clear how the friendship
network is made up of two fairly distinct subgroups, based on
gender. A quickly produced network graphic like this can often
reveal the most important structural patterns contained in the social
network.

## Basic Description

Tukey's original five-number summary was intended to describe the
most important distributional characteristics of a variable,
including its central tendency and variability, using
easy to produce statistical summaries. Similarly, using only
a few functions and lines of R code, we can produce a network
five-number summary that tells us how \emph{large} the network
is, how \emph{densely} connected it is, whether the network is made
up of one or more distinct \emph{groups}, how \emph{compact} it is,
and how \emph{clustered} are the network members.

### Size

The most basic characteristic of a network is its size. The size
is simply the number of members, usually called nodes, vertices or actors.
The \texttt{gorder()} function is the easiest way to get this.
The basic summary of an \texttt{igraph} network object also provides
this information, among other things. The Moreno network has
33 members, based on the \texttt{gorder()} function results and summary calls.


```{r summnetsize}
gorder(Moreno)
summary(Moreno)
```

### Density

Of all the basic characteristics of a social network, density is
among the most important as well as being one of the easiest to
understand. Density is the proportion of observed ties (also
called edges, arcs, or relations) in a network to the maximum
number of possible ties. Thus, density is a ratio that can range
from 0 to 1. The closer to 1 the density is, the more interconnected
is the network.

Density is relatively easy to calculate, although the underlying
equation differs based on whether the network ties are directed
or undirected (see Chapter \ref{chap:directed} for more information
on directed networks). An undirected tie is one with no direction. Collaboration
would be a good example of an undirected tie; if A collaborates with
B, then by necessity B is also collaborating with A. Directed ties,
on the other hand, have direction. Money flow is a good example of
a directed tie. Just because A gives money to B, does not necessarily
mean that B reciprocates. For a directed network, the maximum
number of possible ties among \emph{k} actors is $k*(k-1)$, so the
formula for density is:

\begin{displaymath} \frac{L}{k \times (k-1)} \end{displaymath},

where \emph{L} is the number of observed ties in the network. Density, as
defined here, does not allow for ties between a particular node
and itself (called a loop).

For an undirected network the maximum number of ties is
$k*(k-1)/2$ because non-directed ties should only be counted
once for every dyad (\emph{i.e.,} pair of nodes). So, density for
an undirected network becomes:

\begin{displaymath} \frac{2L}{k \times (k-1)} \end{displaymath}.

The information obtained in the previous section told us that
the Moreno network has 33 nodes and 46 non-directed edges. We
could then use R to calculate that by hand, but it is easier
to simply use the \texttt{edge\_density()} function.

```{r summdensity}
den_hand <- 2*46/(33*32)
den_hand
edge_density(Moreno)
```

### Components

A social network is sometimes split into various subgroups. Chapter
\ref{chap:subgroups} will describe how to use R to identify a wide
variety of network groups and communities. However, a very basic
type of subgroup in a network is a component. An informal definition
of a component is a subgroup in which all actors are connected, directly or indirectly. 
The number of components in a network can be obtained
with the \texttt{count\_components()} function. The more general
function \texttt{components} returns more information about the subgroups,
including the membership, size, and number of separate components.
(Note that the meaning
of components is more complicated for directed networks. See
`?components` for more information.)


```{r summcomponents}
count_components(Moreno)
```

### Diameter

Although the overall size of a network may be interesting, a more
useful characteristic of the network is how compact it is, given its
size and degree of interconnectedness (i.e., density). The diameter of a network
is a useful measure of this compactness. A path is the series of
steps required to go from node A to node B in a network. The shortest
path is the shortest number of steps required to connect the two nodes. 
The diameter then for
an entire network is the longest of the shortest paths across all
pairs of nodes. This is a measure of compactness or network efficiency
in that the diameter reflects the 'worst case scenario' for sending
information (or any other resource) across a network. Although social
networks can be very large, they can still have small diameters because
of their density and clustering (see below).

The only complicating factor for examining the diameter of a network
is that it is formally undefined for networks that contain more than one component.
A typical approach when there are multiple components is to examine
the diameter of the largest component in the network. For the
Moreno network there are two components (see Figure \ref{fig:plotmoreno}).
The smaller component only has 2 nodes. Therefore, we will want to use the
larger component that contains the other 31 connected students. 

Fortunately, the \texttt{diameter()} function in \texttt{igraph} automatically
takes care of this and reports the diameter for the largest component
in a multicomponent network. A diameter
of 11 suggests that this network is not very compact. It takes 11 steps
to connect the two nodes that are situated the furthest apart in this
friendship network.


```{r summdiameter}
diameter(Moreno)
```

### Clustering coefficient


One of the fundamental characteristics of social networks (compared
to random networks) is the presence of clustering, or the tendency
to formed closed triangles. The process of closure occurs in a social
network when two people who share a common friend also become friends
themselves. This can be measured in a social network by examining
its transitivity. Transitivity is defined as the proportion of closed
triangles (triads where all three ties are observed) to the total
number of open and closed triangles (triads where either two or all
three ties are observed). Thus, like density, transitivity is a ratio
that can range from 0 to 1. Transitivity of a network can be calculated
using the \texttt{transitivity()} function. The transitivity for the
4th graders is .29, suggesting a moderate level of clustering
in the classroom network.


```{r summtrans}
transitivity(Moreno)
```


In the rest of this book, we will examine in more detail how the
power of R can be harnessed to explore and study the characteristics
of social networks. The preceding examples show that basic plots
and statistics can be easily obtained. The meaning of these statistics
will always rest on the theories and hypotheses that the analyst
brings to the task, as well as history and experience doing network
analysis with other similar types of social networks.

```{r fivedetachigraph,results='hide'}
detach("package:igraph", unload=TRUE)
```


